---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

<!-- **Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland. -->

**A1** The County that saw the most voters since July 21st was Montgomery County with 4865 registrants. Montgomery County has the highest population of people in Maryland at 1,063,881. SO there is no suprise that they had the highest amount of people register since July 21st.

```{r}
new_voters <- md_voters_2024 |> 
  filter(COUNTY_REGISTRATION_DATE > as.Date("2024-07-21"))

new_voters_by_county <- new_voters |> 
  group_by(COUNTY) |> 
  summarise(COUNTY_REGISTRATION_DATE = n()) |> 
  arrange(desc(COUNTY_REGISTRATION_DATE)) 

print(new_voters_by_county)

```
**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** I used chatGPT for this question because I have been stuck for days and could not figure it out.Compared to the other weeks there is a higher average of registered voters from July 21-August 7th. More poeple registered to vote during this time and that is what the data shows

```{r}
md_voters_2024 <- md_voters_2024 |> 
  mutate(week_number = week(COUNTY_REGISTRATION_DATE)) 

weekly_registrations <- md_voters_2024 |> 
  group_by(week_number) |> 
  summarise(registrant_count = n()) |> 
  arrange(week_number)

print(weekly_registrations)

highlighted_weeks <- unique(week(seq(as.Date("2024-07-21"), as.Date("2024-08-07"), by="day")))


print(weekly_summary)

```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** 
The month that saw the largest number of registations is July and yes that just goes to show that there was an increase in voter registration during july.
```{r}
monthly_registrations <- md_voters_2024 |> 
  mutate(registration_month = floor_date(COUNTY_REGISTRATION_DATE, "month")) |>
  group_by(registration_month) |> 
  summarise(registrant_count = n()) |> 
  arrange(desc(registrant_count))
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** I used chat GPT for this question to help with the outline. The Capitol Region accounts for the most new voters, Democrats, and the total was	19098.

```{r}
library(dplyr)

md_voters_2024 <- md_voters_2024 %>%
  mutate(region = case_when(
    COUNTY %in% c("Allegany", "Garrett", "Washington","Frederick") ~ "Western Maryland",
    COUNTY %in% c("Montgomery", "Prince George's") ~ "Capital Region",
    COUNTY %in% c("Howard", "Baltimore County","Anne Arundel","Carroll") ~ "Central Region",
    COUNTY %in% c("Calvert", "Charles", "Saint Mary's") ~ "Southern Region",
    COUNTY %in% c("Talbot", "Wicomico", "Dorchester","Worcester","Kent","Caroline","Cecil","Queen Anne's","Harford","Somerset") ~ "Eastern Shore",
    COUNTY == "Baltimore City" ~ "Baltimore City",
    TRUE ~ "Other"
  ))
region_party_voters <- md_voters_2024 %>%
  group_by(region,PARTY) %>%
  summarise(new_voters = n()) %>%
  arrange(desc(new_voters))

print(region_party_voters)
```

**Q5** What's an additional question you could ask of this data?

**A5** An interesting question that could be asked about the data is there any party that was more likely to register after July 21st.
