---
title: "lab_01"
author: "Ryan Little"
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Import the Tidyverse. Write the code needed to load the library. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
library(tidyverse)
```

## Load Data

In this lab, we will be recreating the first data story I ever did as a data journalism student at Merrill while working in the Howard Center for Investigative Journalism. First, we need to import the court charges data from Okaloosa County, Florida.

**Task** Using the example from the pre_lab, write code to load the CSV file in the data file and assign it to a good variable name. This data is only charges for homeless people.

```{r}
read_csv("data/okaloosa-court-charges.csv")
```

## Examine the data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.

**Q1** How many rows and columns does this dataframe have?
#There are 15,386 rows and 56 columns 
**A1**

```{r}
glimpse(okaloosa_court_charges)

```

## Answer questions

**Q2** Which charges are homeless people most often charged with?

**A2** # Homeless people are most charged with 893.147(1)

```{r}
okaloosa_court_charges |>
group_by(arrest_charge) |>
summarise(
arrest_charge_count = n()
) |>
  arrange(desc(arrest_charge_count))

```

**Q3** What charges against homeless people have been charged the least?

**A3** # There is a number of arrest charges that are 0

```{r}
okaloosa_court_charges |>
group_by(arrest_charge) |>
summarise(
arrest_charge_count = n()
) |>
  arrange(arrest_charge_count)
```

**Q4** Why do you think there are so many charges identified in **A3** that are tied for the least common? Hint: You can't answer this question with code.

**A4** #I think there are multiple ones tied for A3 because these charges one are just outlining charges and aren't usally charged to the homeless people in this data set

**Q5** Which prosecutor has prosecuted the most charges against homeless people? Is the top prosecutor related to any famous Floridians?

**A5** # ginger bowden madden	has the most charges against homless people. Yes she is Bobby Bowden daughter

```{r}
okaloosa_court_charges |>
group_by(prosecutor) |>

summarise(
prosecutor_charge_count = n()
) |>
  arrange(desc(prosecutor_charge_count))
```


**Q6** Which judge has overseen the most cases involving homeless people? Hint: Cases and charges are not the same thing.

**A6** # Patricia s grinsted	2421	

```{r}
okaloosa_court_charges |>
  group_by(judge) |>
  summarise(
    cases = n_distinct(case_number)
    ) |>
    arrange(desc(cases))
  
```

**Q7** What prosecutor has prosecuted the most cases against a single homeless person?

**A7** #ginger bowden madden	628

```{r}
okaloosa_court_charges |>
group_by(prosecutor) |>
summarise(
case_number= n_distinct(uniform_case_number)
) |>
arrange(desc(case_number))

```

**Q8** We need to create a dataframe that will help us when we are reporting on the ground. Building off our work so far, we need a dataframe that includes the name of every homeless person who was charged with a crime and some important counts about their interactions with the criminal justice system. Create a dataframe that includes the name of each homeless person, the number of charges they have faced, the number of cases they have been involved in, how many dollars they have been assessed in fines, the average fine imposed on them and the median fine imposed upon them. Then sort the data frame so the person with the most charges is at the top. 

**A8**

```{r}
okaloosa_court_charges |>
group_by(defendant) |>
summarise(
prosecutor_charge_count = n(),
mean_balance = mean(balance),
case_number= n_distinct(uniform_case_number),
median_balance = median(balance),
total_assessed = sum(balance),
) |>

arrange(desc(prosecutor_charge_count))



```

**Q9** Please describe a short reporting plan for how you would use the data you have generated in **A8** to report on the criminal justice system in Okaloosa County, Florida. Who are the people we should talk to? What more should we do with our data analysis to be certain we know what the story is?

**A9** # I would want to sit down with some of the homeless individuals who have been charged the most but also have racked up the most in fines. I think it would be helpful to talk to the district attorney and the police commissioner to see if they have any actions in order for homeless people so they won't have to keep committing frivolous fraud.